---
source: Rmd
title: "Introduction to Bioconductor"
teaching: XX
exercises: XX
questions:
- "What is Bioconductor?"
objectives:
- "Install required packages."
keypoints:
- "Bioconductor is a package repository, but also much more than that."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

```{r, include=FALSE}
options(htmltools.dir.version = FALSE)
library(RefManageR)
bib <- ReadBib("../bibliography.bib")
BibOptions(
  longnamesfirst = FALSE,
  check.entries = FALSE,
  bib.style = "authoryear",
  cite.style = "authoryear",
  max.names = 1,
  style = "html",
  hyperlink = "to.bib",
  dashed = TRUE)
```

# What is Bioconductor?

> ## Contribute!
>
> - Stress the many different aspects of the project - not just a package repository.
> - Bullet point below + A short paragraph for details and links.
>
{: .callout}

- A scientific project
- A package repository
- A support site
- A Slack workspace
- A mailing list for developers
- Conferences, courses and workshops
- A community

## A scientific project

The original publication describes the aims and methods of the project at its inception `r Citet(bib, "Gentleman2004")`.


## A package repository

[BiocViews][biocviews-software] are a predefined - and evolving - set of labels used to categorize all the packages available in the Bioconductor project.


## A support site

The Bioconductor [support site][bioc-support-site] provides a platform where users and developers can communicate freely (following the Bioconductor [Code of Conduct][bioc-code-of-conduct]) to discuss issues on a range of subjects, ranging from packages to conceptual questions about best practices.

## A Slack workspace

The Bioconductor [Slack workspace][bioc-slack] is an open space that all community members are welcome to join (for free) and use for rapid interactions.
Currently, the "Pro" pricing plan kindly supported by core funding provides:

- Unlimited message archive
- Unlimited apps
- Group video calls with screen sharing
- Work securely with other organizations using Slack Connect

A wide range of channels have been created to discuss a range of subjects, and community members can freely join the discussion on those channels of create new ones to discuss new subjects.

Important announcements are posted on the `#general` channel.

> ## Note
>
> Users are encouraged to use the Bioconductor [support site][bioc-support-site] to raise issues that relevant to the wider community.
> The Slack workspace is often most useful for live discussions, and widely subscribed channels (e.g. `#general`) should be used with moderation.
>
{: .callout}


## A mailing list for developers

The <bioc-devel@r-project.org> mailing list is used for communication between package developers, and announcements from the Biocondutor core team.


## Conferences, courses and workshops

The Bioconductor community regularly organizes a number of events throughout the year and across the world.
For example:

- The annual BioC summer conference in North America
- Regional conference in winter (e.g. BioC Europe, BioC Asia)
- Summer schools (e.g., CSAMA)
- Online meetings open to all community members (e.g., Bioconductor Developers Forum)

[Course materials][bioc-course-materials] are regularly uploaded on the Bioconductor website following each of those events.


## A community

> ## Contribute!
>
> - Add some details to the role of each panel below.
>
{: .callout}

- [Scientific Advisory Board][scientific-advisory-board]
- [Technical Advisory Board][technical-advisory-board]
- [Community Advisory Board][community-advisory-board]


# Installing packages

```{r, eval=FALSE}
install.packages("BiocManager")
BiocManager::install("BiocPkgTools")
BiocManager::version()
BiocManager::valid()
```

* Release cycle - release vs devel


# Explore the package universe

```{r}
library(BiocPkgTools)
```


# References

```{r, results='asis', echo=FALSE}
PrintBibliography(bib)
```

[biocviews-software]: https://www.bioconductor.org/packages/release/BiocViews.html#___Software
[bioc-code-of-conduct]: https://www.bioconductor.org/about/code-of-conduct/
[bioc-slack]: https://bioc-community.herokuapp.com/
[bioc-course-materials]: https://bioconductor.org/help/course-materials/
[scientific-advisory-board]: https://bioconductor.org/about/scientific-advisory-board/
[technical-advisory-board]: https://bioconductor.org/about/technical-advisory-board/
[community-advisory-board]: https://bioconductor.org/about/community-advisory-board/
