---
source: Rmd
title: "Installing Bioconductor"
teaching: XX
exercises: XX
questions:
- "How do I install Bioconductor packages?"
objectives:
- "Install BiocManager."
- "Install Bioconductor packages."
keypoints:
- "BiocManager is used to install Bioconductor packages (but also from CRAN and GitHub), and check for updates."
- "BiocManager safely manages packages from the incremental releases of Bioconductor."
- "The BiocManager package is available from the CRAN repository."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
stopifnot(require(BiocStyle))
```

# Installing BiocManager

The `r BiocStyle::Biocpkg("BiocManager")` package is the entry point into Bioconductor.
This is technically the only Bioconductor package distributed on the CRAN repository.
As such, the package can be installed using the traditional `install.packages()` function.

```{r, eval=FALSE}
install.packages("BiocManager")
```


# Bioconductor releases and current version

The Bioconductor project produces two releases each year, one around April and another one around October.

The April release of Bioconductor coincides with the annual release of R.
Packages in that Bioconductor release are tested for the upcoming version of R.
Users must install the new version of R to access the new version of those packages.

The October release of Bioconductor continue to work with the same version of R for that annual cycle.

Each time a new release is made, the minor version of all the packages in the Bioconductor repository is incremented by one.
In particular, the version of the `r BiocStyle::Biocpkg("BiocVersion")` package is used to represent the version of the Bioconductor project.

Once the `r BiocStyle::Biocpkg("BiocManager")` package is installed, the `BiocManager::version()` function displays the version (i.e., release) of the Bioconductor project that is currently active in the R session.

```{r, message=FALSE}
BiocManager::version()
```

It is possible to specify a version of Bioconductor, which in turn will install the latest version of Bioconductor packages for that particular Bioconductor release.
For instance:

```{r, echo=FALSE, results='asis'}
version <- BiocManager::version()
cmd <- sprintf('BiocManager::install(version = "%s")', as.character(version))
cat(
    "~~~",
    cmd,
    "~~~",
    "{: .language-r}",
    sep = "\n")
```

Note that `BiocManager::install(version = ...)` can be used to update a library of Bioconductor packages to a certain version of Bioconductor, either a more recent or an older release.

> ## Release cycle - release and devel branches
>
> ### Release branches
>
> Bioconductor uses the [Git][git-website] version control system to manage its package repository.
> For each new Bioconductor release, a new branch is created in the Git repository; those are referred to as _release_ branches.
> Release branches allow users to install stable versions of packages that were tested together for a given version of Bioconductor, when we specificy `BiocManager::install(version = ...)`
>
> Development on the _release_ branches is restricted.
> Older _release_ branches are entirely frozen, meaning that no further update is allowed on those branches.
> When users request a package from release branches, they receive the latest version of the package **for that release**.
> 
> Only the latest release branch allows updates from package maintainers, but those are restricted to critical bug fixes.
> This means that for each 6-month release cycle, users can expect packages on the latest branch to be reasonably stable.
>
> ### Devel branches
>
> Meanwhile, the main branch of the Git repository (historically called `master`) is referred to as the _devel_ branch.
>
> The _devel_ branch allow developers to continue updating the packages frequently, without affecting users or disrupting workflows.
> Typically, the _devel_ branch is mainly used by other developers and the Bioconductor build system, to run tests using the latest code of every package in the Bioconductor repository.
> However, users can also access packages on the _devel_ branch using `BiocManager::install(version = ...)` with `version` set to one minor version greater than the latest Bioconductor _release_ version (e.g. if release is `3.13`, then devel is `3.14`).
>
> ### Transition between devel and release - the release process
>
> After a new release branch is created, the minor version of every single package on the _devel_ branch is incremented by one.
> This includes the `r BiocStyle::Biocpkg("BiocVersion")` package, which marks the value of the next version of Bioconductor.
>
{: .callout}


# Check for updates

The `BiocManager::valid()` function checks the version of currently installed packages, and checks whether a new version is available for any of them on the Bioconductor repository.

Conveniently, if any package can be updated, the function generates and displays the command needed to update those packages.
Users simply need to copy-paste and run that command in their R console.

If everything is up-to-date, the function will simply print `TRUE`.

```{r, message=FALSE}
BiocManager::valid()
```


# Installing packages

The `BiocManager::install()` function is used to install packages.
The function first searches for the requested package(s) on the Bioconductor repository, but falls back on the CRAN repository and also supports installation from GitHub repositories.

For instance, we can install the `r BiocStyle::Biocpkg("BiocPkgTools")` package:

```{r, eval=FALSE}
BiocManager::install("BiocPkgTools")
```


# Explore the package universe

```{r, message=FALSE}
library(BiocPkgTools)
```

> ## Contribute!
>
> - Demonstrate relevant functions of the `r BiocStyle::Biocpkg("BiocPkgTools")` package.
>
{: .callout}

# Finding a suitable package

* biocViews

[git-website]: https://git-scm.com/
