---
source: Rmd
title: "Installing Bioconductor"
teaching: XX
exercises: XX
questions:
- "How do I install Bioconductor packages?"
objectives:
- "Install BiocManager."
- "Install Bioconductor packages."
keypoints:
- "The BiocManager package is available from the CRAN repository."
- "BiocManager is used to install Bioconductor packages (but also from CRAN and GitHub), and check for updates."
- "BiocManager safely manages packages from the incremental releases of Bioconductor."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
```

# Installing BiocManager

The `r BiocStyle::Biocpkg("BiocManager")` is the entry point into Bioconductor.
This is technically the only Bioconductor package distributed on the CRAN repository.
As such, the package can be installed using the traditional `install.packages()` function.

```{r, eval=FALSE}
install.packages("BiocManager")
```

# Bioconductor releases and current version

The Bioconductor project produces two releases each year, typically one around April and another one towards October.

The April release of Bioconductor coincides with the annual release of R.
This means that the new release of packages was tested for that new version of R, and requires user to install that new version of R to access the new version of those packages.

The October release of Bioconductor will continue to work with the same version of R.

The `BiocManager::version()` function displays the version (i.e., release) of Bioconductor that is currently active in the R session.

```{r}
BiocManager::version()
```

# Title

```{r}
BiocManager::valid()
```

# Installing packages

The `BiocManager::install()` function is used to install packages.
The function first searches for the requested package(s) on the Bioconductor repository, but falls back on the CRAN repository and also supports installation from GitHub repositories.

For instance, we can install the `r BiocStyle::Biocpkg("BiocPkgTools")` package:

```{r, eval=FALSE}
BiocManager::install("BiocPkgTools")
```

* Release cycle - release vs devel

# Explore the package universe

```{r, message=FALSE}
library(BiocPkgTools)
```

> ## Contribute!
>
> - Demonstrate relevant functions of the `r BiocStyle::Biocpkg("BiocPkgTools")` package.
>
{: .callout}

# Finding a suitable package

* biocViews
