---
source: Rmd
title: "Introduction and setup"
teaching: XX
exercises: XX
questions:
- "Am I using the correct version of R for this lesson?"
- "Why does my version of R matter?"
objectives:
- "Ensure that participants are using the correct version of R to reproduce exactly the contents of this lesson."
keypoints:
- "Participants will only be able to install the version of Bioconductor packages described in this lesson and reproduce their exact outputs if they use the correct version of R."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

This lesson was developed and tested with `r R.version.string`.

Take a moment to launch RStudio and verify that you are using R version `4.1.x`, with `x` being any patch version, e.g. `4.1.2`.

```{r}
R.version.string
```

This is important because Bioconductor uses the version of R running in the current session to determine the version of Bioconductor packages that can be installed in the R library associated with the current R session.
Using a different version of R while following this lesson may lead to unexpected results.
