---
source: Rmd
title: "Standard Bioconductor classes"
teaching: XX
exercises: XX
questions:
- "What is the S4 class system?"
- "How do we access information stored in S4 objects?"
- "How to we combine methods and classes from different packages into a practical workflow?"
objectives:
- "Describe the S4 class system."
- "Identify common classes and methods used in the Bioconductor project."
- "Create objects and process them using methods from differerent packages in a single workflow."
keypoints:
- "Common methods and classes promote interoperability between packages, reduce redundant efforts by package developers, and facilitate the writing of workflows"
- "S4 classes can represent arbitrarily complex data structures."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
```

# Install packages

```{r, eval=FALSE}
install.packages("BiocManager")
BiocManager::install()
```

# Why standard classes?

# S4 class system

# Standard classes

> ## Contribute!
> 
> - Add bullet points below, and for each point add a short paragraph for details and links.
>
{: .callout}

* Biostrings
* IRanges and GRanges
* GAlignments
* SummarizedExperiment and SingleCellExperiment
* MultiAssayExperiment
* QFeatures


# Biostrings

> ## Contribute!
> 
> Add content about:
>
> - Which package defines the classes.
> - What the classes represent.
> - Examples of packages and methods that can process objects from those classes.
>
{: .callout}


# IRanges and GRanges

> ## Contribute!
> 
> Add content about:
>
> - Which package defines the classes.
> - What the classes represent.
> - Examples of packages and methods that can process objects from those classes.
>
{: .callout}


# GAlignments

> ## Contribute!
> 
> Add content about:
>
> - Which package defines the classes.
> - What the classes represent.
> - Examples of packages and methods that can process objects from those classes.
>
{: .callout}


# SummarizedExperiment and SingleCellExperiment

> ## Contribute!
> 
> Add content about:
>
> - Which package defines the classes.
> - What the classes represent.
> - Examples of packages and methods that can process objects from those classes.
>
{: .callout}


# MultiAssayExperiment

> ## Contribute!
> 
> Add content about:
>
> - Which package defines the classes.
> - What the classes represent.
> - Examples of packages and methods that can process objects from those classes.
>
{: .callout}


# QFeatures

> ## Contribute!
> 
> Add content about:
>
> - Which package defines the classes.
> - What the classes represent.
> - Examples of packages and methods that can process objects from those classes.
>
{: .callout}
