---
source: Rmd
title: Working with gene annotations
teaching: XX
exercises: XX
---

---

```{r, echo=FALSE, purl=FALSE, message=FALSE}
source("download_data.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Explain how gene annotations are managed in the Bioconductor project.
- Identify Bioconductor packages and methods available to fetch and use gene annotations.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What Bioconductor packages provides methods to efficiently fetch and use gene annotations?
- How can I use gene annotation packages to convert between different gene identifiers?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
```

```{r, include=FALSE}
options(htmltools.dir.version = FALSE)
library(RefManageR)
library(bibtex)
bib <- ReadBib("files/bibliography.bib")
```

## Install packages

Before we can proceed into the following sections, we install some Bioconductor
packages that we will need.
First, we check that the `r BiocStyle::Biocpkg("BiocManager")` package is
installed before trying to use it; otherwise we install it.
Then we use the `BiocManager::install()` function to install the necessary
packages.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("biomaRt", "org.Hs.eg.db"))
```

## Overview

Packages dedicated to query gene annotations exist in the 'Software' and
'Annotation' categories of the Bioconductor [biocViews][biocviews], according to
their nature.

In the 'Software' section, we find packages that do not actually contain gene
annotations, but rather dynamically _query_ them from online resources
(e.g.,[Ensembl BioMart][biomart-ensembl]).
One such Bioconductor package is `r BiocStyle::Biocpkg("biomaRt")`.

Instead, in the 'Annotation' section, we find packages that do contain
annotations.
Examples include `r BiocStyle::Biocpkg("org.Hs.eg.db")`,
`r BiocStyle::Biocpkg("EnsDb.Hsapiens.v86")`,
and `r BiocStyle::Biocpkg("TxDb.Hsapiens.UCSC.hg38.knownGene")`.

In this episode, we will demonstrate the two approaches:

* Querying annotations from the Ensembl Biomart API using the `r BiocStyle::Biocpkg("biomaRt")` package.
* Querying annotations from the `r BiocStyle::Biocpkg("org.Hs.eg.db")` annotation package.

## Querying annotations from Ensembl BioMart

### Pros and cons

Pros:

* Automatically access the latest information
* Minimal storage footprint on the user's computer

Cons:

* Requires a live and stable internet connection throughout the analysis.
* Reproducibility may not be possible if the resource is updated without access
  to archives.
* Data may be organised differently in each resource.
* Custom code may be needed to access and retrieve data from each resource.

### The Ensembl BioMart

[Ensembl BioMart][biomart-ensembl] is a robust data mining tool designed to
facilitate access to the vast array of biological data available through the
Ensembl project.

The [BioMart web interface][biomart-ensembl] enables researchers to efficiently
query and retrieve data on genes, proteins, and other genomic features
across multiple species.
It allows users to filter, sort, and export data based on various attributes
such as gene IDs, chromosomal locations, and functional annotations.

### The Bioconductor `biomaRt` package

`r BiocStyle::Biocpkg("biomaRt")` is a Bioconductor software package that
enables retrieval of large amounts of data from Ensembl BioMart tables
directly from an R session where those annotations can be used.

Let us first load the package:

```{r, message=FALSE, warning=FALSE}
library(biomaRt)
```

### Available marts

Ensembl BioMart organises its diverse biological information into four databases
also known as *marts* or *biomarts*.
Each mart focuses on a different type of data.

Users must select the mart corresponds to the type of data they are interested
in before they can query any information from it.

The function `listMarts()` can be used to display the names of those marts.
This is convenient as users do not need to memorise the name of the marts,
and the function will also return an updated list of names if any mart is
renamed, added, or removed.

```{r}
listMarts()
```

In this demonstration, we will use the biomart called `ENSEMBL_MART_ENSEMBL`,
which contains the Ensembl gene set.

Notably, the `version` columns also indicates the version of the biomart.
The Ensembl BioMart is updated regularly (multiple times per year).
By default, `r BiocStyle::Biocpkg("biomaRt")` functions access the latest
version of each biomart.
This is not ideal for reproducibility.

Thankfully, Ensembl BioMart archives past versions of its mars in a way that
is accessible both programmatically, and on its website.

The function `listEnsemblArchives()` can be used to display all the versions of
Ensembl Biomart accessible.

```{r}
listEnsemblArchives()
```

In the output above, the key piece of information is the `url` column, which
provides the URL that `r BiocStyle::Biocpkg("biomaRt")` functions will need to
access data from the corresponding snapshot of the Ensembl BioMart.

### Connecting to a biomart

The two pieces of information collected above -- the name of a biomart
and the URL of a snapshot -- is all that is needed to connect to a BioMart
database reproducibly.

The function `useMart()` can then be used to create a connection.
The connection is traditionally stored in an object called `mart`.

```{r}
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "https://may2024.archive.ensembl.org")
```

[biocviews]: https://www.bioconductor.org/packages/release/BiocViews.html
[biomart-ensembl]: https://www.ensembl.org/biomart/martview
