---
source: Rmd
title: Working with gene annotations
teaching: XX
exercises: XX
---

---

```{r, echo=FALSE, purl=FALSE, message=FALSE}
source("download_data.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Explain how gene annotations are managed in the Bioconductor project.
- Identify Bioconductor packages and methods available to fetch and use gene annotations.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What Bioconductor packages provides methods to efficiently fetch and use gene annotations?
- How can I use gene annotation packages to convert between different gene identifiers?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
```

```{r, include=FALSE}
options(htmltools.dir.version = FALSE)
library(RefManageR)
library(bibtex)
bib <- ReadBib("files/bibliography.bib")
```

## Install packages

Before we can proceed into the following sections, we install some Bioconductor
packages that we will need.
First, we check that the `r BiocStyle::Biocpkg("BiocManager")` package is
installed before trying to use it; otherwise we install it.
Then we use the `BiocManager::install()` function to install the necessary
packages.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("biomaRt", "org.Hs.eg.db", ""))
```

## Overview

Packages dedicated to query gene annotations exist in the 'Software' and
'Annotation' categories of the Bioconductor [biocViews][biocviews], according to
their nature.

In the 'Software' section, we find packages that do not _contain_ gene
annotations,
but rather dynamically _query_ them from online resources
(e.g.,[Ensembl BioMart][biomart-ensembl]).
One such Bioconductor package is `r BiocStyle::Biocpkg("biomaRt")`.

Instead, in the 'Annotation' section, we find packages that _contain_
annotations.
Examples include `r BiocStyle::Biocpkg("org.Hs.eg.db")`,
`r BiocStyle::Biocpkg("EnsDb.Hsapiens.v86")`,
and `r BiocStyle::Biocpkg("TxDb.Hsapiens.UCSC.hg38.knownGene")`.


[biocviews]: https://www.bioconductor.org/packages/release/BiocViews.html
[biomart-ensembl]: https://www.ensembl.org/biomart/martview
